Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-18 paper/main.tex
rewrite amir subsection
- cover new force type
- explain observed profiles
- discuss model reduction

We model the force exerted by the fluid flow onto the individual vertices by
\begin{equation}
    \vec{F}_{i,\text{external}} = C |\vec{x}_i - \vec{x}_{i+1}|(y-y_b)\sin(\beta_i)\vec{e}_x
\end{equation}
which is proportional to the length of the rod segment projected along the y-axis.
It acts similarly to Stoke's Law~\cite{Stokes1856} along the axis of the fluid flow (x-axis).
Here $\beta_i$ is the angle of the rod edge between $\vec{x}_{i+1}-\vec{x}_i$ and the x-axis.
Amir et. al~\cite{Amir2014_2} calculated the flow profile for this particular setup.
Since the bacterium does not reach the peak of the velocity profile, we can estimate the slope via a
linear dependence on the pertruding segment along the y-axis~\cite{Wolschin2021}.
In particular, this enforces the correct boundary values the flow profile and the resulting force
$\vec{F}_{i,\text{external}}|_{y=0}=0$.

We extract the positions of the rod from the data images (see
Figure~\ref{fig:amir-bending-simulation} (A)) and compare the simulated positions with the extracted
ones using the direct comparison algorithm (see Subsection~\ref{subsec:direct-comparison}).
Our simulation is split into two phases; one with an external force that initially bends the rod,
and a second one during which the force is disabled and the cell returns to its original shape.
We assume that these time intervals are identically long since the data from the movies does not
allow for a finer resolution of the dynamics.

Since this model is void of any cell-cell interactions or cell division, the dynamics of the cell
can be described by a combination of growth and physical mechanics (see
subsection~\ref{subsection:mechanical-model-mechanics}).
We are thus left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$ (see table~\ref{table:list-of-parameters}).
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (orange) and final
(blue) position of the rod using the optimized parameters.
While the parameters for the dragging force, growth rate and damping can be identified, the same
can not be said for the rod rigidity and spring tension, which both are not identifiable.
However this behaviour is expected since, the shape of the rod is almost exclusively straight, thus
adding no further information except that a lower threshold of rigidity and spring tension should be
present.

Furthermore the used images only contain snapshots of a fully-bent or fully-relaxed rod without any
intermediate steps.
This effectively limits us to 2 datapoints and is likely the key factor that leads to the two local
minima observed in the damping profile.
This is further supported by the observation that the contributions to the cost function is evenly
distributed between the bent and relaxed position of our system.

\textbf{TODO what is the parameter $s$? Explain more clearly.}\\
To remove this non-identifiability we reduce the number of parameters by setting $\gamma=\gamma_0/s$
and $\eta=\eta_0 s$ for the rod rigidity and spring tension, resp.
We then optimize for this parameter and pick $\gamma_0=\SI{15}{\micro\metre}$ with
$\eta_0=\SI{150}{\micro\metre\per\second\squared}$ which corresponds to $60\%$ of the previously considered
interval.
\textbf{Plots are not in supplement anymore}
The corresponding plots can be seen in \textbf{TODO}.
\textbf{TODO Make this more clear. Transient information would lead to more information. We only
have initial and final snapshots.}

>>>>

We model the force exerted by the fluid flow onto the individual vertices by
\begin{equation}
    \vec{F}_{i,\text{external}} = \frac{C}{2} |\vec{x}_i - \vec{x}_{i+1}|(yR - y^2)\sin(\beta_i)\vec{e}_x
\end{equation}
which is proportional to the length of the rod segment projected along the y-axis.
Similarly to Stoke's Law~\cite{Stokes1856}, it acts along the axis of the fluid flow (x-axis).
Here $\beta_i$ is the angle of the rod edge between $\vec{x}_{i+1}-\vec{x}_i$ and the x-axis.
The flow profile for this particular setup is proportional to $(2yR-y^2)$ where $y$ is the vertical
displacement measured from the end of the tube~\cite{Wolschin2021}
(see Figure~\ref{fig:amir-bending-simulation}) and $R$ is the total size of the chamber along the
y-axis.
Since the bacterium does not reach the peak of the velocity profile, we can estimate the slope via a
linear dependence on the pertruding segment along the y-axis~\cite{Wolschin2021}.
In particular, this enforces the correct boundary values the flow profile and the resulting force
$\vec{F}_{i,\text{external}}|_{y=0}=0$.

We extract the positions of the rod from the data images (see
Figure~\ref{fig:amir-bending-simulation} (A)) and compare the simulated positions with the extracted
ones using the direct comparison algorithm (see Subsection~\ref{subsec:direct-comparison}).
Our simulation is split into two phases; One with the external flow force that initially bends the
rod and lasts $\SI{7}{\second}$ and a second phase during which the force is disabled and the cell
returns to its original shape over a span of $\SI{17}{\second}$.

Since this setup is void of any cell-cell interactions or cell division, the dynamics of the cell
can be described by a combination of growth and physical mechanics (see
subsection~\ref{subsection:mechanical-model-mechanics}).
We are thus left with 5 parameters:
Magnitude of the drag force $C$, growth rate $\alpha$, rigidity of the rod $\eta$, spring tension
$\gamma$ and damping constant $\lambda$ (see table~\ref{table:list-of-parameters}).
Figure~\ref{fig:amir-bending-simulation} (B) shows a snapshot of the initial (upright, light blue)
and intermediate (bent, dark blue) position of the rod using the optimized parameters.
The likelihood profiles displayed in figure~\ref{fig:amir-bending-simulation} (C-D, blue curves)
show that each parameter is identifiable from the given data.
However when it comes to the damping constant $\lambda$, its overall numerical value and steepness
of the curve compared to the results for the other parameters are much lower.
We can assess if the system could be described by a reduced model by setting $\lambda=0$, thus
effectively removing any damping.
The curves which are obtained in this way are displayed in orange and show that the system is also
fully identifiable in its reduced version.

It should be noted that the used images only contain snapshots of a fully-bent or fully-relaxed rod
without any intermediate steps.
This effectively limits us to effectively 2 datapoints and is likely the key factor that leads to
the previously described phenomenon.
The damping force is of dynamic nature, and takes no effect in the two steady states of the rod
being either fully bent or fully relaxed.
Gathering additional transient information would allow us to better estimate the damping parameter.
